


## Data structures
Main reference here is @Gro2014 and @Wic2017.

### Vectors {#sec-r-rb-ds-vectors}
Vector is one of the basic data structure in R. It is created by `c()` function. Sometimes it is called `atomic vector`. You may store any data types in it. R recognizes six basic types: *double*, *integers*, *characters*, *logicals*, *complex* and *raw*.

The data type inside a vector can be checked by `typeof` function.

```{r}
die <- c(1, 2, 3, 4, 5, 6)
typeof(die)
```
For consecutive numbers, an easier way to create vector is to use `:`.

```{r}
die <- 1:6
```


::: {.callout-caution}
Note that vector index starts from 1 in R, while list index starts from 0 in Python.
:::

```{r}
die[1]
```
When slicing with vectors, don't forget to use `c()`.

```{r}
die[c(2, 3)]
```

```{r}
die[2:3]
```

You may use `length()` function to get its length.

```{r}
length(die)
```



### Attributes

R objects may have attributes. Attributes won't be shown by default when you show the object. You may find the attributes of a R object by calling the `attributes()` function.

The following example show that the vector `die` defined in @sec-r-rb-ds-vectors doesn't have attributes.

```{r}
attributes(die)
```

Attributes can be read and write using `attr` function. See the following example.

::: {#exm-}
```{r}
attr(die, 'date') <- '2022-01-01'
die
attr(die, 'date') <- NULL
die
```
:::

You may think attributes as metadata attached to a R object. They are used to tell some useful infomation of the object. Some functions will interact with certain attributes. R itself treat attributes `class`, `comment`, `dim`, `dimnames`, `names`, `row.names` and `tsp` specially. We will only talk about `class` and `names` here. `dim` will be discussed in the next section. Others will be discussed when we use them.

1. `class`: This is different from the class in Python. `class` in R is an attribute which talks about the class of an object. If the attribute `class` is not assigned to an object, the object will have an implicit class: `matrix`, `array`, `function`, `numeric` or the result of `typeof`. 

`attr(x, 'class')` will show the "external" class of an object. You may also use `class(x)` to read and write attribute `class`. If the `class` is not assigned, `class(x)` will show the implicit class, while `attr(x, 'class')` will show `NULL`. 


::: {#exm-}

```{r}
attr(die, 'class')
class(die)
class(die) <- 'a die'
attr(die, 'class')
class(die)
```
:::
```{r}
#| echo: false
class(die) <- NULL
```

2. `names`: This attribute is used to name each element in a vector. After the names are assigned, it won't be displayed below the data like other attributes. It will be displayed above the data with correct alignment. Similar to `class`, you may use `names()` to read and write the attribute.


::: {#exm-}
```{r}
names(die) <- c('one', 'two', 'three', 'four', 'five', 'six')
die
attributes(die)
names(die)
is.vector(die)
```
:::


::: {.callout-tip}
When you store different types of data into a single vector in R, R will convert them into a single type. The default way to do so is

- if there are only logicals and numbers, logicals will be converted to numbers by `TRUE->1` and `FALSE->0`.
- if characters are presented, all are converted to characters by what it is.


```{r}
c(1, TRUE)
c('1', 1, TRUE)
```
:::



::: {.callout-note}
We can apply regular operators to vectors. The defaul way is to apply the operators element-wise.
:::


### matrices and arrays


```{r}
m <- matrix(c(1,2,3,4,5,6), nrow=2)
m[1, ]
```
A matrix has `dim` attribute.

```{r}
dim(m)
```

Note that by assigning and removing `dim` attribute, you may change the object between vectors and matrices.


::: {#exm-}

```{r}
m
is.matrix(m)
is.vector(m)
dim(m)
dim(m) <- NULL
m
is.matrix(m)
is.vector(m)
```
:::


::: {.callout-note}
The `dim` of a matrix/vector can be a long vector. In this case, it will become an array. 
:::

### factors
Factor is speical vector. It is a way to handle categorical data. The idea is the limit the possible values. In a factor all possible values are called `level`, which is an attribute. 


::: {#exm-}
We would like to talk about all months. We first define a vector of the valid levels:

```{r}
month_levels <- c(
  "Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec"
)
```
Then we could start to transform some month vector into factors, by the function `factor()`.

```{r}
x1 <- c("Dec", "Apr", "Jan", "Mar")
y1 <- factor(x1, level=month_levels)
sort(x1)
sort(y1)
```
Note that sorting `y1` is based on the levels.

```{r}
x2 <- c("Dec", "Apr", "Jam", "Mar")
y2 <- factor(x2, level=month_levels)
y2
```
Note that `y2` contains `NA` value since there is an entry in `x2` that is not valid.
:::



### Lists
List is very similar to a vector. The main difference is that vector can only store values, while list can store objects. The most typical example of objects is another vector. Please see the following example.

::: {#exm-}

```{r}
c(1:2, 3:4)
list(1:2, 3:4)
```

:::

::: {.callout-note}
The attributes of an object is stored in an array.

```{r}
m <- matrix(c(1,2,3,4,5,6), nrow=2)
a <- attributes(m)
class(a)
```

:::



### `data.frame`
Data.Frame is a list with the `class` attribute `data.frame`, together with some restriction on the shape of each columns. You may think about it in terms of tables.

```{r}
df <- data.frame(face = c("ace", "two", "six"),
                 suit = c("clubs", "clubs", "clubs"),
                 value = c(1, 2, 3))
df
```

- Data Frame group vectors. Each vector represents a column.
- Different column can contain a different type of data, but every cell within one column must be the same type of data.
- `data.frame()` can be used to create a data.frame. 
- The type of a data.frame is a list. Similar to matrix comparing to vector, a `data.frame` is a `list` with `class` `data.frame`, as well as a few other attributes.



### Examples


::: {#exm-}
Consider a date.frame representing a deck of cards. Here we use `expand.grid()` to perform the Cartesian product.

```{r}
suit <- c('spades', 'hearts', 'clubs', 'diamonds')
face <- 1:13
deck <- expand.grid(suit, face)
head(deck)
```
We may assign names to change the column names.

```{r}
names(deck) <- c('suit', 'face')
head(deck)
```

Note that since `suit` and `face` are two vectors, `merge()` can also do the Cartesian product. `expand.grid()` is good for both vectors and data.frame.


```{r}
deck <- merge(suit, face)
head(deck)
```

:::


### Load data

#### build-in datasets
R has many build-in datasets. You may use `data()` to see all of them. Here are a few common datasets.

- `mtcars`: Motor Trend Car Road Tests: The data was extracted from the 1974 Motor Trend US magazine, and comprises fuel consumption and 10 aspects of automobile design and performance for 32 automobiles (1973–74 models)

```{r}
data(mtcars)
```

- `iris`: iris data set gives the measurements in centimeters of the variables sepal length, sepal width, petal length and petal width, respectively, for 50 flowers from each of 3 species of iris. The species are Iris setosa, versicolor, and virginica.

```{r}
data(iris)
```

- `ToothGrowth`: ToothGrowth data set contains the result from an experiment studying the effect of vitamin C on tooth growth in 60 Guinea pigs. 

```{r}
data(ToothGrowth)
```

- `PlantGrowth`: Results obtained from an experiment to compare yields (as measured by dried weight of plants) obtained under a control and two different treatment condition.

```{r}
data(PlantGrowth)
```

- `USArrests`: This data set contains statistics about violent crime rates by us state.

```{r}
data(USArrests)
```

#### Read from files
The build-in `read.csv()` function can directly read `.csv` file into a data.frame.


::: {#exm-}
We use the file `yob1880.txt` from Chapter 5 here. Put the file in the working folder and run the following code.

```{r}
#| eval: false
df <- read.csv('yob1880.txt', header = FALSE)
head(df)
```

```{r}
#| echo: false
df <- read.csv('assests/datasets/yob1880.txt', header = FALSE) 
```
We may also manually assign columns names.

```{r}
names(df) <- c('name', 'sex', 'counts')
head(df)
```
:::


::: {.callout-note}
To save data is straightforward. 

```{r}
#| eval: false
write.csv(df, file='df.csv', row.names=FALSE)
```
:::





### Flow control

#### `for` loop


::: {#exm-}

```{r}
for (x in 1:10){
    print(x)
}
```

:::


#### `if-else` 


::: {#exm-}

```{r}
a <- 200
b <- 33

if (b > a) {
  print("b is greater than a")
} else if (a == b) {
  print("a and b are equal")
} else {
  print("a is greater than b")
}
```

:::


#### Functions
The standard format to define a function is `my_function <- function(input) {}` where the function name is on the left side of `<-`, the input arguments are in the `()`, and the function body is in `{}`. The output of the last line of the function body is the return value of the function.

::: {#exm-}

```{r}
myfunction <- function() {
    die <- 1:6
    sum(die)
}

myfunction()
```
If you just type the function name without `()`, R will return the definition of the function.

```{r}
myfunction
```

:::



::: {.callout-tip}
The function `sample(x)`: `sample` takes a sample of the specified size from the elements of `x` using either with or without replacement. 

`sample(x, size, replace = FALSE, prob = NULL)`: 

- `x`: either a vector of one or more elements from which to choose, or a positive integer.
- `size`: a non-negative integer giving the number of items to choose.
- `replace`: should sampling be with replacement?
- `prob`: a vector of probability weights for obtaining the elements of the vector being sampled.
:::


