## Modifying values

### Changing values in place

You can use R’s notation system to modify values within an R object. 

- In general when working with vectors, the two vectors should have the same length.
- If the lengths are different, R will repeat the shorter one to make it match with the longer one. This is called the *vector recycling* rule. R will throw a warning if the two lengths are not proposional.

::: {#exm-}
```{r}
#| error: true
1:4 + 1:2
1:4 + 1:3
```
:::



- We may create values that do not yet exist in the object. R will expand the object to accommodate the new values.


::: {#exm-}

```{r}
vec <- 1:6
vec
vec[7] <- 0
vec
```

:::


::: {#exm-}

```{r}
df <- data.frame(a=c(1,2), b=c('a', 'b'))
df
df$c <- 3:4
df
```

:::


### Logical subsetting
We could compare two vectors element-wise, and the result is a logical vector. Then we could use this result to subset the vector / data.frame.



::: {#exm-}

```{r}
suit <- c('spades', 'hearts', 'clubs', 'diamonds')
face <- 1:5
deck <- expand.grid(suit, face)
```

```{r}
deck$Var1 == 'hearts'
deck$Var2[deck$Var1 == 'hearts']
deck[deck$Var1 == 'hearts',]
```
We could directly assign values to the subset. Note that the following assignment create a new column with `NA` values.


```{r}
deck$Var3[deck$Var1 == 'hearts'] <- 1
deck
```


:::


:::{.callout-tip}
Other than the regualr logical operators, R provides a speical one: `%in%`. 

`x %in% y`: Is `x` in the vector `y`?

If `x` is a vector, the output is a vector with the same length as `x`, telling whether each element of `x` is in `y` or not.
:::


::: {.callout-tip}
Other than the regular Boolean operators, R provides two special ones: `any` and `all`. 

- `any(cond1, cond2, ...)`: Are any of these conditions true?
- `all(cond1, cond2, ...)`: Are all of these conditions true?
:::


### Missing values `NA`

In R, missing values are `NA`, and you can directly work with `NA`. Any computations related to `NA` will return `NA`.

1. `na.rm`: Most R functions come with the optional argument `na.rm`. If you set it to be `TRUE`, the function will ignore `NA` when evaluating the function.


::: {#exm-}

```{r}
mean(c(NA, 1:50))
mean(c(NA, 1:50), na.rm=TRUE)
```

:::

2. `is.na()`: This is a function testing whether an object is `NA`.